---
title: "CourseProject VitaC and ToothGrowth"
output: pdf_document
---

##INTRODUCTION

####On Vitamin C and ToothGrowth 

For the 2nd part of the project, we are assigned to investigate the ToothGrowth dataset provided in R. As a brief background of the dataset, the data was drawn from a study regarding the effect of vitamin C. The length of odontoblasts in 60 guinea pigs were recorded after treatment of varying vitamin C dosage. Dosages (ranging from 0.5 to 2mg/day) were also delivered by two methods, via orange juice or ascorbic acid.

##DATA OVERVIEW
```{r, echo=FALSE, warning=FALSE}
#Loading Dataset and Libraries
data(ToothGrowth)
library(dplyr)
library(ggplot2)
library(gridExtra)
```

```{r, echo=FALSE}
#Quick overview of data structure and features
str(ToothGrowth)
head(ToothGrowth, 5)
tail(ToothGrowth, 5)
```

##EXPLORATORY AND DATA SUMMARY -- Univariate

```{r, echo=FALSE}
#Summary Statistics of Tooth Length
summary(ToothGrowth$len)
#Distribution of Tooth Length
a <- ggplot(ToothGrowth, aes(x=len)) +
  geom_density() +
  scale_x_continuous(breaks=seq(0, 35,5)) +
  xlab("Tooth Length")
b <- ggplot(ToothGrowth, aes(x=factor(1), y=len), color="red") +
  geom_boxplot() +
  scale_y_continuous(breaks=seq(0,35,5)) +
  xlab("Tooth Length") + ylab("Unit Measure")
grid.arrange(a, b, ncol=2, widths=c(1,0.25), heights=c(7,1))
```

```{r, echo=FALSE}
#Summary Statistics of Tooth Length
summary(ToothGrowth$len)
#Supp(Delivery Method) and Dosage Count
a <- ggplot(ToothGrowth, aes(x=supp)) +
  geom_bar(aes(fill=factor(dose))) +
  scale_y_discrete(breaks=seq(0,30,5))
b <- ggplot(ToothGrowth, aes(x=factor(dose))) +
  geom_bar() +
  facet_grid(.~supp)
grid.arrange(a, b, ncol=2, widths=c(1,1), heights=c(7,1))
```
From this graphic, we are able to pull additional information regarding the study. Of the 60 guinea pigs, 30 each are assigned to each delivery method (left graphic). Of each delivery method, 10 of each are assigned to the three dosage categories (0.5, 1 and 2)

##EXPLORATORY AND DATA SUMMARY -- Bivariate

####Dosage

```{r, echo=FALSE}
#Boxplot and Scatterplot - Length by Dose
a <- ggplot(ToothGrowth, aes(x=factor(dose), y=len)) +
  geom_boxplot(aes(stat="identity")) +
  scale_y_continuous(breaks=seq(0,35,5))
b <- ggplot(ToothGrowth, aes(x=dose, y=len)) +
  geom_point(aes(color=factor(supp))) +
  scale_y_continuous(breaks=seq(0,35,5))
grid.arrange(a, b, ncol=2, widths=c(0.5,1), heights=c(7,1))
#Density Distribution of Length by Group(Dose)
ggplot(ToothGrowth, aes(x=len, fill=supp)) +
  geom_density(alpha=0.2)
```

Visually comparing the medians, we can observe that guinea pigs that recieved higher dosage (1-2mmg) tend to have higher tooth length compared to those that only have 0.5mmg dosage. Further, when adding delivery method represented by color (left graph), we see that for 0.5 and 1.0 dosage, guinea pigs tended to have higher tooth length.

```{r, echo=FALSE}
#Statistics -- Length by Dose
ToothGrowth %>% 
  group_by(supp) %>% 
  summarise(n= length(supp), GrpMean=round(mean(len),2), GrpVar=round(var(len),2), GrpSD=round(sd(len),2))
```


###Supp (Delivery Method)

```{r, echo=FALSE}
#Boxplot and Scatterplot -- Length by Supp(Delivery)
a <- ggplot(ToothGrowth, aes(x=supp, y=len)) +
  geom_boxplot(aes(stat="identity")) +
  scale_y_continuous(breaks=seq(0,35,5))
b <- ggplot(ToothGrowth, aes(x=supp, y=len)) +
  geom_point(aes(color=factor(dose))) +
  scale_y_continuous(breaks=seq(0,35,5))
grid.arrange(a, b, ncol=2, widths=c(0.5,1), heights=c(7,1))
#Density Distribution of Length by Supp
ggplot(ToothGrowth, aes(x=len, fill=factor(dose))) +
  geom_density(alpha=0.2)
```

filler

```{r, echo=FALSE}
#Statistics -- Length by Supp(Delivery)
ToothGrowth %>% 
  group_by(factor(dose)) %>% 
  summarise(n= length(dose), GrpMean=round(mean(len),2), GrpVar=round(var(len),2), GrpSD=round(sd(len),2))
```


##STATISTICAL TESTS

##CONCLUSION