---
title: "Course Project - Exp Distributions"
output: pdf_document
---

##Introduction to the Course Project
Under Coursera and John Hopkin's University's Statistical Inference course, the course project requests that we investigate exponential distributions and verifiy its mean and variance metric's behavior to that prescribed by the Central Limit Theorum. 

To invetigate this, we will first demonstrate the therorhetical mean, standard deviations and variance estimates provided by the formula given exponential distribtion with $\lambda = 0.2$. Following, we will utilize 40 simulation each with $n=1000$ to test the formula's accuracy. 

##Central Limit Theorum and Exponential Distributions
The central limit theorum states that:
$$\lim{n \to infty}, sqrt(n)*(S_n - \mu)/\sigma = N(0, 1)$$

In other words, as $n$ increases, the distribution of difference between the sample sampling statistic (mean or variance) and its population parameter will approximate the normal distribution with mean 0 and variance of $\sigma^2$. 

The applications of this can be useful as we will demonstrate as this feature is consistent for many type of distributions that are not inherently Gaussian.

##Theoretical Estimates of Mean, Variance and Standard Deviation
In an expoenential distribution with $\lambda = 0.2$,
$mu$ = $E(X)$ = 1/$\lambda$ = 1/0.2 = 5
$Var(X)$ = 1/$\lambda^2$ = 1/0.04 = 25

```{r, echo=TRUE}
MeanEST <- 1/0.2
VarEST <- 1/(0.2^2)
```


##Simulated Mean, Variance and Standard Deviation
```{r, echo=TRUE}
n <- 1
dat <- c()
while (n <=1000){
  dat <- rbind(dat, rexp(40, rate=0.2))
  n <- n + 1
}

zMean <- function(x, n){sqrt(n)*(mean(x) - MeanEST)/VarEST}
zVar <- function(x, n){sqrt(n)*(var(x) - VarEST)/VarEST}

out <- data.frame(
  EXmean = apply(dat, 1, zMean, n),
  EXvar = apply(dat, 1, zVar, n)
)
```


```{r, echo=TRUE}
library(ggplot2)
library(gridExtra)
ggplot(out, aes(x = EXmean)) +
  geom_density()

ggplot(out, aes(x = EXvar)) +
  geom_density()

```

##Part II: ToothGrowth
For the 2nd part of the project, we are assigned to investigate the ToothGrowth dataset provided in R. As a brief background of the dataset, the data was drawn from a study regarding the effect of vitamin C. The length of odontoblasts in 60 guinea pigs were recorded after treatment of varying vitamin C dosage. Dosages (ranging from 0.5 to 2mg/day) were also delivered by two methods, via orange juice or ascorbic acid.

##Data Overview
```{r, echo=TRUE}
str(ToothGrowth)
```

##Exploratory -- Univariate

```{r, echo=TRUE}
#Summary Statistics of Tooth Length
summary(ToothGrowth$len)
#Distribution of Tooth Length
a <- ggplot(ToothGrowth, aes(x=len)) +
  geom_density() +
  scale_x_continuous(breaks=seq(0, 35,5)) +
  xlab("Tooth Length")
b <- ggplot(ToothGrowth, aes(x=factor(1), y=len), color="red") +
  geom_boxplot() +
  scale_y_continuous(breaks=seq(0,35,5)) +
  xlab("Tooth Length") + ylab("Unit Measure")
grid.arrange(a, b, ncol=2, widths=c(1,0.25), heights=c(7,1))
```

```{r, echo=TRUE}
#Summary Statistics of Tooth Length
summary(ToothGrowth$len)
#Boxplot and Scatterplot -- Length by Supp
a <- ggplot(ToothGrowth, aes(x=supp)) +
  geom_bar(aes(fill=factor(dose))) +
  scale_y_discrete(breaks=seq(0,30,5))
b <- ggplot(ToothGrowth, aes(x=factor(dose))) +
  geom_bar() +
  facet_grid(.~supp)
grid.arrange(a, b, ncol=2, widths=c(1,1), heights=c(7,1))
#Density Distribution of Length by Supp
ggplot(ToothGrowth, aes(x=len, fill=factor(dose))) +
  geom_density(alpha=0.2)
```
From this graphic, we are able to pull additional information regarding the study. Of the 60 guinea pigs, 30 each are assigned to each delivery method (left graphic). Of each delivery method, 10 of each are assigned to the three dosage categories (0.5, 1 and 2)

##Exploratory -- Bivariate
```{r, echo=TRUE}
#Boxplot and Scatterplot - Length by Dose
a <- ggplot(ToothGrowth, aes(x=factor(dose), y=len)) +
  geom_boxplot(aes(stat="identity")) +
  scale_y_continuous(breaks=seq(0,35,5))
b <- ggplot(ToothGrowth, aes(x=dose, y=len)) +
  geom_point(aes(color=factor(supp))) +
  scale_y_continuous(breaks=seq(0,35,5))
grid.arrange(a, b, ncol=2, widths=c(0.5,1), heights=c(7,1))
#Density Distribution of Length by Group(Dose)
ggplot(ToothGrowth, aes(x=len, fill=supp)) +
  geom_density(alpha=0.2)
```
Visually comparing the medians, we can observe that guinea pigs that recieved higher dosage (1-2mmg) tend to have higher tooth length compared to those that only have 0.5mmg dosage. Further, when adding delivery method represented by color (left graph), we see that for 0.5 and 1.0 dosage, guinea pigs tended to have higher tooth length.

```{r, echo=TRUE}
#Boxplot and Scatterplot -- Length by Supp(Delivery)
a <- ggplot(ToothGrowth, aes(x=supp, y=len)) +
  geom_boxplot(aes(stat="identity")) +
  scale_y_continuous(breaks=seq(0,35,5))
b <- ggplot(ToothGrowth, aes(x=supp, y=len)) +
  geom_point(aes(color=factor(dose))) +
  scale_y_continuous(breaks=seq(0,35,5))
grid.arrange(a, b, ncol=2, widths=c(0.5,1), heights=c(7,1))
```

##Data Summary
```{r, echo=TRUE}
#Statistics -- Length by Dose
ToothGrowth %>% 
  group_by(supp) %>% 
  summarise(n= length(supp), GrpMean=round(mean(len),2), GrpVar=round(var(len),2), GrpSD=round(sd(len),2))
```


```{r, echo=TRUE}
#Statistics -- Length by Supp(Delivery)
ToothGrowth %>% 
  group_by(factor(dose)) %>% 
  summarise(n= length(dose), GrpMean=round(mean(len),2), GrpVar=round(var(len),2), GrpSD=round(sd(len),2))
```
