---
title: "lc2 - On Intercept and Coefficients"
author: "Rock Chi"
date: "Sunday, October 18, 2015"
output: html_document
---

##Linear Regression
The basic linear model:
$$ Y_i = \beta_0 + \beta_1 X_i + \epsilon_{i} $$

Moving Parts of the Model:
$\epsilon_{i}$ are assumed iid $N(0, \sigma^2)$

- cumulation of errors from variables missing from the model which we should have modeled


$E[Y_i ~|~ X_i = x_i] = \mu_i = \beta_0 + \beta_1 x_i$

- the expected value of response Y given regressor x is estimated as the value on the line, $\beta_0 + \beta_1 x_i$


$Var(Y_i ~|~ X_i = x_i) = \sigma^2$

- the variance of response Y at any particular point is $\sigma^2$
- this is different from the variance of response Y. It is the variance of Y around the regression line at any point. Thus, lower than the actual variance of Y because some of the variance has been already explained the regressor


##Understanding Regression Coefficients
Intercept Interpretation
$$ E[Y | X = 0] = \beta_0 + \beta_1 \times 0 = \beta_0 $$

- $\beta_0$, the intercept, is the expected value of the response when predictors are held at 0
- sometimes, predictors held at 0 may not be of interest (ie. heart pressure)

Shifting Regression Variables Away from Intercept 0
 $$ Y_i = \beta_0 + \beta_1 X_i + \epsilon_i = \beta_0 + a \beta_1 + \beta_1 (X_i - a) + \epsilon_i = \tilde \beta_0 + \beta_1 (X_i - a) + \epsilon_i $$

- $X$ values are all shifted by a constant of $a$. Intercept is changed, but slope is not
- $a \beta_1$ is added back in changing the intercept
- a possible constant that could be subtracted is the mean
 
###Simulated Example of Mean Centering
From the Diamond Data, we estimate diamond prices (Singapore Dollars) with carat.
```{r, echo=TRUE}
library(UsingR)
data(diamond)

fit1 <- lm(price ~ carat, data=diamond)
coef(fit1)
```

From this output, we see when diamond carat is 0, price will be -259! This is not useful for interpretation and inference. Thus, we use mean centering to tease out better information.
```{r, echo=TRUE}
fit2 <- lm(price ~ I(carat-mean(carat)), data=diamond)
coef(fit2)
```

This output expresses the intercept as the expected price of average sized diamonds.

##Coefficient Interpretation
- $\beta_1$ is the expected change in response for a 1 unit change in the predictor

On changing scale/unit of X
$X$. $$ Y_i = \beta_0 + \beta_1 X_i + \epsilon_i = \beta_0 + \frac{\beta_1}{a} (X_i a) + \epsilon_i = \beta_0 + \tilde \beta_1 (X_i a) + \epsilon_i $$

- From this equation, we understand that by multiplying regressor $x$ by $a$, we are dividing the coefficients by $a$

###Simulated Example of Unit Changing
Using the diamond dataset again, we predict diamond prices using the same model. However, instead of price change per carat, we would like to see price change per 0.1 carat.

```{r, echo=TRUE}
fit3 <- lm(price ~ I(carat*10), data=diamond)
coef(fit3)
```


##Summary of Scaling and Unit Change
To illustrate the changes from the three transformations:
```{r, echo=TRUE}
ggplot(diamond, aes(carat, y=price)) +
        stat_smooth(method="lm", formula=y ~ x, se=FALSE, color="red") +
        stat_smooth(method="lm", formula=y ~ I(x-mean(x)), se=FALSE, color="blue") +
        stat_smooth(dmethod="lm", formula=y ~ I(x*10), se=FALSE, color="green")
```

Red line is the non-transformed regression line. X-Intercept is at a negative value.

Blue line is the mean-shifted regression line. X-Intercept is now at the expected value of average price for average size

Green line is the unit shifted regression line. As we can see, slope is changed. This is because unit scale of slope has been changed.


###Predict Function
The following demonstrates how the predict function work.
```{r, echo=TRUE}
newx <- c(0.16, 0.27, 0.34)
#Using the formula (coef 1 == intercept/beta0, coef 2 == beta1)
coef(fit1)[1] + coef(fit1)[2] * newx
#Using prediction function
predict(fit1, newdata=data.frame(carat=newx))
```
As we can see, both are equivalent.